Modelling BCG vaccination in the UK

### 2019-07-29 rmds/08-model-development.Rmd
cleanded out cache

\begin{equation}
\lambda^k_a = \frac{\beta_a}{N} \sum\limits_{i=1}^{A} \rho_i C_{ai}\left(\frac{M\iota_i}{\nu_a^P} + \sum\limits_{j = u,v}P^j_i\right)
  (\#eq:force-of-infection)
\end{equation}

Where $\iota_a$ is the age stratified number of non-UK born pulmonary cases notified in a given year, $P^{u,v}_a$ is the number of vaccinated, and unvaccinated, pulmonary TB cases, $\rho_a$ is the age-specific proportion of cases that are smear positive, $\nu^P_a$ is the age-specific rate of starting treatment for active pulmonary TB, $C_{ai}$ is the age-stratified contact matrix (Section \@ref(contact-matrix)), $\beta_a$ is the age-stratified transmission probability, and $M$ is the mixing rate between the UK born and non-UK born population.

@@

```{r disease-model}
disease_params_table %>% 
  select(Parameter, Description, Distribution, Units, Method, Type) %>% 
kable(
  caption.short = "Dynamic disease model parameters, descriptions, prior distributions, units, method used to derive the prior distribution and the type.", 
  caption = "Dynamic disease model parameters, descriptions, prior distributions, units, method used to derive the prior distribution and the type (i.e data derived, literature, assumption). All data based parameters are included. P = pulmonary TB, E = extra-pulmonary TB, v = vaccinated, i = age at vaccination, $\\mathcal{U}$ = Uniform, $\\mathcal{N}$ = Normal", booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header", "hold_position")) %>% 
  column_spec(c(5), width = "6cm") %>% 
  column_spec(c(2), width = "4cm") %>%
  column_spec(c(3), width = "6cm") %>%
  column_spec(c(1,4,6), width = "1.5cm") %>% 
  landscape()
```

@@

\begin{equation}
\text{Non UK-born scaling} = \frac{e^{\left(\frac{t - 1960}{\text{ln}(2) * \iota_{\text{scale}}}\right)} - 1}{e^{\left(\frac{2000 - 1960}{\text{ln}(2) * \iota_{\text{scale}}}\right)} - 1}
  (\#eq:non-uk-born-scale)
\end{equation}

@@

\begin{equation}
C^{\text{scaled}}_{\text{eff}} = \begin{cases}
C^{\text{hist}}_{\text{eff}} & \text{if } t < 1950, \\
C_{\text{eff}} + \left(C^{\text{hist}}_{\text{eff}} -  C_{\text{eff}} \right) \left(1 - \frac{t-1950}{C^{\text{year}}_{\text{eff}}-1950}\right) & \text{if } 1950 \leq t \leq 1990, \\
C_{\text{eff}} & \text{if } t > 1990. \end{cases}
  (\#eq:scaled-contacts)
\end{equation}

@@

Vynnycky et al. found that the effective contact rate (i.e the transmission probability times the contact rate) for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since 1990 (Chapter \@ref(data)) and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is normally distributed with a mean of 1 and a standard deviation of 0.5. Similarly, for the historic effective contact rate I have assumed a normal distribution between with a mean of 10 and a standard deviation of 2.5. I have also assumed that the historic contact rate declines linearly over time, starting in 1950, reducing to 0 by $C^{\text{year}}_{\text{eff}}$. The prior for this year has been assumed to be normally distributed with a mean of 1985, and a standard deviation of 5 years. It has been truncated so that the lower bound is 1975 and the upper bound is 1995. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation between children and adults ($\beta_{\text{child}}$) and between children, adults and older adults ($\beta_{\text{child}}$ and $\beta_{\text{older-adults}}$) also being considered. The priors for these modifiers are normally distributed around 1 with a standard deviation of 0.5 (truncated to be greater than 0). The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

Model simulations were begun in 1931 due to the availability of population data from the 1931 census and because data on live births was only available from 1929. The demographic model is initialised using the age grouped 1931 census data, with the assumption that the entirety of the population is UK born. Initially it is assumed that there is no BCG vaccination and no TB treatment. TB treatment is assumed to begin in 1952 with the discovery of isoniazid and BCG vaccination begins at school-age (15 years old) in 1953. BCG vaccination coverage is assumed to vary randomly over the time horizon of the model but to have the same distribution at all time points. The assumed distribution is normal with a mean of 80% and a standard deviation of 5%. The duration with active TB is assumed to decrease from the introduction of treatment in 1952 through to 1990 when it is assumed that detection rates were equivalent to those seen today.

@@

The model is initialised with the number of pulmonary and extra-pulmonary cases reported in 1931. The high risk latent population is initialised by scaling the number of observed cases in 1931 by the proportion of high risk latent cases that develop active TB, the duration that these cases are high risk and then dividing by the infectious period. The low risk latent population is then initialised by scaling the high risk latent population by the cumulative sum of the age distribution of UK born cases in 2000. Finally, the initial susceptible population is based on the population estimate from the 1931 census minus the assumed initial latent cases.

@@

In the next chapter, I will consider the evidence for modifying the transmission probability by age using three distinct scenarios (Table \@ref(tab:summary-scenarios)). Identifying if the transmission probablity varies with age is important as it may alter the distribution and number of TB cases. This would impact the observed effects of the BCG vaccine and is therefore of primary importance.

@@

The semi-stochastic transmission dynamic model of TB transmission and BCG vaccination presented in this chapter provides a detailed overview of the features required to reproduce the observed epidemiology of TB in England. The model was robustly parameterised using routine surveillance data where available and otherwise using literature sources. The assumptions required by the model can be explored by fitting the model to observed data and assessing the goodness of fit. This is the focus of the next chapter. In addition the model may also be used to explore the impact of current and historic BCG vaccination policy, both in the observed data and projected into the future. This is explored in the final two chapters of this thesis.

>>>>

\begin{equation}
\lambda^k_a = \frac{\beta_a}{N} \sum\limits_{i=1}^{A} \rho_i C_{ai}\left(\frac{M_{a}\iota_i}{\nu_a^P} + \sum\limits_{j = u,v}P^j_i\right)
  (\#eq:force-of-infection)
\end{equation}

Where $\iota_a$ is the age stratified number of non-UK born pulmonary cases notified in a given year, $P^{u,v}_a$ is the number of vaccinated, and unvaccinated, pulmonary TB cases, $\rho_a$ is the age-specific proportion of cases that are smear positive, $\nu^P_a$ is the age-specific rate of starting treatment for active pulmonary TB, $C_{ai}$ is the age-stratified contact matrix (Section \@ref(contact-matrix)), $\beta_a$ is the age-stratified transmission probability, and $M_a$ is the age-stratified mixing rate between the UK born and non-UK born population.

@@

```{r disease-model, eval = FALSE}
disease_params_table %>% 
  select(Parameter, Description, Distribution, Units, Method, Type) %>% 
kable(
  caption.short = "Dynamic disease model parameters, descriptions, prior distributions, units, method used to derive the prior distribution and the type.", 
  caption = "Dynamic disease model parameters, descriptions, prior distributions, units, method used to derive the prior distribution and the type (i.e data derived, literature, assumption). All data based parameters are included. P = pulmonary TB, E = extra-pulmonary TB, v = vaccinated, i = age at vaccination, $\\mathcal{U}$ = Uniform, $\\mathcal{N}$ = Normal", booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header", "hold_position")) %>% 
  column_spec(c(5), width = "6cm") %>% 
  column_spec(c(2), width = "4cm") %>%
  column_spec(c(3), width = "6cm") %>%
  column_spec(c(1,4,6), width = "1.5cm") %>% 
  landscape()
```

@@

\begin{equation}
\text{Non UK-born scaling} = \frac{e^{\left(\frac{t - 1960}{\text{ln}(2) * (-1) * \iota_{\text{scale}}}\right)} - 1}{e^{\left(\frac{2000 - 1960}{\text{ln}(2) * (-1) * \iota_{\text{scale}}}\right)} - 1}
  (\#eq:non-uk-born-scale)
\end{equation}

@@

\begin{equation}
C^{\text{scaled}}_{\text{eff}} = \begin{cases}
C^{\text{hist}}_{\text{eff}} & \text{if } t < 1935, \\
C_{\text{eff}} + \left(C^{\text{hist}}_{\text{eff}} -  C_{\text{eff}} \right) \left(\frac{t-1935}{C^{\text{half-life}}_{\text{eff}}}\right)^{1/2} & \text{if } 1935 \leq t \leq 1980, \\
C_{\text{eff}} & \text{if } t > 1980. \end{cases}
  (\#eq:scaled-contacts)
\end{equation}

@@

Vynnycky et al. found that the effective contact rate (i.e the transmission probability times the contact rate) for TB was approximately 22 in 1900 and fell to approximately 1 in 1990.[@Vynnycky1999] Incidence rates have increased since the early 1980's (Chapter \@ref(data)) and it is unclear what impact this has had on the effective contact rate. I have assumed that the effective contact rate is normally distributed with a mean of 1 and a standard deviation of 0.5. For the historic effective contact rate I have assumed a uniform distribution with a lower bound of the current effective contact rate and an upper bound of 20. I have also assumed that the historic contact rate declines over time, starting in 1935, reducing to the current effective contact rate in 1980. The speed of this decay is set by defining a half-life ($C^{\text{half-life}}_{\text{eff}}$). The prior for the half-life is assumed to be normally distributed with a mean of 5 years and a standard deviation of 5 (truncated to be greater than 0). This prior is based on the observed trend in notifications. Additional age stratification of $\beta$ is explored by including modifiers for certain age-groups. The baseline scenario is that no modification is required, with variation explored for young adults (15-29; $\beta_{\text{young adult}}$) as a scenario. The priors for this modifier was assumed to be uniform, bounded by 0 and 10. The contact rate is estimated by averaging the total age-specific contact rates estimated from POLYMOD data (Section \@ref(contact-matrix)) on an annual basis.

@@

Model simulations were begun in 1931 due to the availability of population data from the 1931 census and because data on live births was only available from 1929. The demographic model is initialised using the age grouped 1931 census data, with the assumption that the entirety of the population is UK born. Initially it is assumed that there is no BCG vaccination and recovery from active TB takes 2 years. TB treatment is assumed to begin in 1952 with the discovery of isoniazid and BCG vaccination begins at school-age (15 years old) in 1953. BCG vaccination coverage is assumed to vary randomly over the time horizon of the model but to have the same distribution at all time points. The assumed distribution is normal with a mean of 75% and a standard deviation of 5%. The duration with active TB is assumed to decrease from the introduction of treatment in 1952 through to 1990 when it is assumed that detection rates were equivalent to those seen today.

@@

The model is initialised with the number of pulmonary and extra-pulmonary cases reported in 1931. The high risk latent population is initialised by scaling the number of observed cases in 1931 by the proportion of high risk latent cases that develop active TB, the duration that these cases are high risk and then dividing by the infectious period. The low risk latent population is then initialised by scaling the high risk latent population by the cumulative sum of the age distribution of UK born cases in 2000, reduced by 50% to account for mortality. Finally, the initial susceptible population is based on the population estimate from the 1931 census minus the assumed initial latent cases.

@@

In the next chapter, I will consider the evidence for modifying the transmission probability, and non-UK born mixing, by age using three distinct scenarios (Table \@ref(tab:summary-scenarios)). Identifying if the transmission probablity, or non-UK born mixing, varies with age is important as it may alter the distribution and number of TB cases. This would impact the observed effects of the BCG vaccine and is therefore of primary importance.

@@

The semi-stochastic transmission dynamic model of TB transmission and BCG vaccination presented in this chapter provides a detailed overview of the features required to reproduce the observed epidemiology of TB in England. The model was robustly parameterised using routine surveillance data where available and otherwise using literature sources. The assumptions required by the model can be explored by fitting the model to observed data and assessing the goodness of fit. This is the focus of the next chapter. In addition the model may also be used to explore the impact of current and historic BCG vaccination policy, both in the observed data and projected into the future. This is explored in the final chapter of this thesis.

### 2019-07-29 rmds/09-model-fitting.Rmd
cleanded out cache

This chapter details an approach to challenging a infectious disease model to data using the state-space model formulation and bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, strengths and limitations of the approach, and areas for further work.

@@

* Mixing rate between UK born and non-UK born ($M$).
* Scaling on non-UK born cases ($\iota_{\text{scale}}$).
* Effective contact rate ($c_{\text{eff}}$).
* The year that the current effective contact rate was reached ($c^{\text{year}}_{\text{eff}}$).
* Historic effective contact rate ($c^{\text{hist}}_{\text{eff}}$).
* Half life of the effective contact rate ($c^{\text{hist}}_{\text{half}}$).
* Systematic reporting error ($E_{\text{syst}}$).
* Reporting noise ($E_{\text{noise}}$).

In addition for scenarios with age variable transmission probabilities the following paramters may also be fitted to,

* Transmission probablity modifier for children ($\beta_{\text{child}}$).
* Transmission probablity modifier for older adults ($\beta_{\text{older-adults}}$).

@@

I have implemented this as a two step process for each fitted scenario. Firstly, I used the Nelder-Mead simplex method, via LibBi,[@Murray2015] to find a parameter set that optimised the maximum likelihood. I then initialised a 1000 step PMCMC chain with this parameter set, using 256 particles in the particle filter. I then used `rbi.helpers`,[@Funk:2019uw] as outlined above, to estimate the number of particles required to produce a log-likelihood variance of less than 1 for this sample of the posterior distribiution, using 250 samples per step and starting with 64 particles. I initially planned to repeat this process for multiple draws from the posterior distribution but this proved to be infeasible given the compute available. A target of 5 for the log-likelihood variance was chosen as a smaller target could not be feasibly achieved given the compute resources available. Additionally 256 was specified as the maximum number of feasible particles to use in the particle filter.

@@

I have implemented PRCC using the `epiR` R package^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/test_sensitivity.R],[@EpiR] using the samples from the posterior distribution of the model calculated during the SMC^2 step. Parameter sensitivity measures such as PRCC must be calculated seperately for each output time point. I calculated the PRCC for each fitted parameter, at the final time point with fitted data (2004), for both overall TB incidence rates and TB incidence rates in children. These results are then summarised by plotting the absolute PRCC values for each output measure, indicating the direction of correlation using colour^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/plot_sensitivity.R].

@@

1. Model initialisation using minimal error checking and single precision computation. Implemented using the `disable-assert` adn `enable-single` flags in LibBi.[@Murray2015] Outputs are only given for times with oberserved data and a subset of parameters are recorded for final reporting^[Model code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/inst/bi/BaseLineModel.bi].

2. 5000 parameter sets were taken from the prior distribution and the model was then simulated for each one.

@@

7. SMC^2 model fitting with 5000 initial parameter particles. Particle rejuvernation was set to trigger when the effective sample size decreased below 1000, with 4 MCMC steps used each time.

@@

10. Parameter sensitivity was estimated by calculating the partial rank correlation coffecient (PRCC) for each model parameter, for the final time point fitted too (2004), for both overall TB incidence rates and TB incidence rates in children. Results were then plotted in order of the absolute magnitude of the correlation, with the direction of the correlation determined using colour. The `epiR` package was used to compute the PRCC.[@EpiR]

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC^2 algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle reuvernation steps. This resulted in spuriously tight priors. Ad-hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and any conclusions drawn from them must be appropriately caveated.

@@

@@

### Model fit to historic TB incidence rates

* Plot of historic TB data + fit from model
* Table of data + model fit + uncertainty
* Comment on model fit

>>>>

This chapter details an approach to challenging a infectious disease model to data using the state-space model formulation and bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad-hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

* Mixing rate between UK born and non-UK born ($M$).
* Scaling on non-UK born cases ($\iota_{\text{scale}}$).
* Effective contact rate ($c_{\text{eff}}$).
* Historic effective contact rate ($c^{\text{hist}}_{\text{eff}}$).
* Half life of the effective contact rate ($c^{\text{hist}}_{\text{half}}$).
* Low risk latent activation rate modifier for older adults (70+) ($\epsilon^{\text{older-adult}}_L$).
* Systematic reporting error ($E_{\text{syst}}$).
* Reporting noise ($E_{\text{noise}}$).

In addition for scenarios with age variable transmission probabilities or non-UK born mixing the following paramters may also be fitted to,

* Transmission probablity modifier for young adults ($\beta_{\text{young adult}}$).
* Non-UK born mixing modifier for young adults ($M_{\text{young adult}}$).

@@

I have implemented this as a two step process for each fitted scenario. Firstly, I used the Nelder-Mead simplex method, via LibBi,[@Murray2015] to find a parameter set that optimised the maximum likelihood. I then initialised a 1000 step PMCMC chain with this parameter set, using 256 particles in the particle filter. I then used `rbi.helpers`,[@Funk:2019uw] as outlined above, to estimate the number of particles required to produce a log-likelihood variance of less than 1 for this sample of the posterior distribiution, using 250 samples per step and starting with 16 particles. I initially planned to repeat this process for multiple draws from the posterior distribution but this proved to be infeasible given the compute available. A target of 5 for the log-likelihood variance was chosen as a smaller target could not be feasibly achieved given the compute resources available. Additionally 256 was specified as the maximum number of feasible particles to use in the particle filter.

@@

I have implemented PRCC using the `epiR` R package^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/test_sensitivity.R],[@EpiR] using the samples from the posterior distribution of the model calculated during the SMC^2 step. Parameter sensitivity measures such as PRCC must be calculated seperately for each output time point. I calculated the PRCC for each fitted parameter, at the final time point with fitted data (2004), for overall TB incidence rates. These results are then summarised by plotting the absolute PRCC values, indicating the direction of correlation using colour^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/plot_sensitivity.R].

@@

1. Model initialisation using minimal error checking and single precision computation. Implemented using the `disable-assert` adn `enable-single` flags in LibBi.[@Murray2015] Outputs are only given for times with observed data and a subset of parameters are recorded for final reporting^[Model code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/inst/bi/BaseLineModel.bi].

2. 2000 parameter sets were taken from the prior distribution and the model was then simulated for each one.

@@

7. SMC^2 model fitting with 2000 initial parameter particles. Particle rejuvernation was set to trigger when the effective sample size decreased below 25%, with 5 MCMC steps used each time.

@@

10. Parameter sensitivity was estimated by calculating the partial rank correlation coffecient (PRCC) for each model parameter, for the final time point fitted too (2004), foroveral l TB incidence rates. Results were then plotted in order of the absolute magnitude of the correlation, with the direction of the correlation determined using colour. The `epiR` package was used to compute the PRCC.[@EpiR]

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC^2 algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvernation steps. This resulted in spuriously tight priors. Ad-hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and any conclusions drawn from them must be appropriately caveated.

@@

### Particle and proposal calibration

After development of the model fitting pipeline but before results could be produced both the `optimise` (from `rbi`) and `adapt_particle` (from `rbi.helpers`) began to error with a difficult to debug `LibBi` error message. This meant that steps 3-5 of the model fitting pipeline could not be used with the final model. As a work around the maximum permitted number of particles (256) was then used for all model fitting (increasing the number of particles was also explored as detailed in the previous section).

As discussed, there were multiple issues with fitting the final model and this made it diffult to determine what the mean of posterior distribution was. This made manually tuning the proposal distribution challenging and so instead a standard deviation of 10% was used. This value was chosen as it increased the acceptance rate by limiting each rejuvernation step to a relatively small subset of the prior distribution whilst not preventing the exploration of new parameter space in scenarios when model fits from the initial particle sample were poor.

@@
